---
title: "MLearning"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

## Project Summary

**Goal:** Predict diabetes status from demographic and clinical risk factors using the NHANES dataset.

**Data:** National Health and Nutrition Examination Survey collected in the US. Variables include age, BMI, direct cholesterol, marital status, and gender.

**Models:** Logistic Regression and Random Forest (tidymodels framework).

**Key Finding:** Random Forest outperformed logistic regression across all metrics. Sensitivity improved from 4% (logistic regression) to 34% (Random Forest), with ROC-AUC of 0.89. Both models struggled with the minority class due to class imbalance.

**Tools:** R, tidymodels, ranger, ggstatsplot, cowplot, gtsummary

## Introduction

In this project we will use the NHANES dataset to predict diabetes given the available risk factors. The National Health ans Nutrition Survey is a a program in the US designed to assess the health and nutritional status pf adults, and children in the US. The data includes demographic, socio-economic, dietary, and health-related information.

### Loading packages including the dataset.

```{r}
#| include: false
library(NHANES)
library(tidymodels)
library(tidyverse)
library(janitor)
library(cowplot)
library(gtsummary)
library(ranger)
library(randomForest)
#install.packages("ggstatsplot")
library(ggstatsplot)
```

### Data Preparation

We are going to save the dataset into the nhanes_df object to maintain the original dataset intact.

```{r Data subsetting}
nhanes_df<-NHANES |> 
  select(Diabetes,DirectChol,BMI,MaritalStatus,Age,Gender) |>
  drop_na() |> # Assuming variables are missing completely at Random
  clean_names()

# Changing the levels for appropriate analysis
nhanes_df<-nhanes_df |> 
  mutate(diabetes=factor(diabetes, 
                         levels = c("Yes", "No"))) |> 
           glimpse()
```

## Train/Test Split

In the code below, we are splitting our data into training and testing sets (0.8, 0.2) and stratify by the diabetes to preserve class proportions.

```{r Resampling}
set.seed(123) #For reproducubility
ml_split<-initial_split(nhanes_df,
                        prop = 0.8,
                        strata = diabetes)

ml_training<-ml_split |> 
  training()

ml_test<-ml_split |>
  testing()
  
```

## Logistic Regression

```{r Model specification}

lr_model<-logistic_reg() |> 
  set_engine("glm") |> 
  set_mode("classification") 

```

### Correlation Check

```{r Checking for Correlated predictors}
ml_training |> 
  select_if(is.numeric) |> 
  ggcorrmat(colors = c("#B2182B", "White", "#4D4D4D"),
            title = "Colleration Matrix"
            )
```

Using the hypothetical threshold of 0.8, we can conclude that the predictors are not collerated. In the code below, we are going to fit both models using the fit function. After which we are going to collect and combine predictions, and load them.

In the code below we are going to specify a recipe object after which we will add steps for engineering our features (feature engineering). The steps are to preprocess the data into a form that will allegedly improve our analysis.

### Workflow and Fit

```{r Recipe Object}
set.seed(123)
lr_recipe<-recipe(diabetes~.,data = ml_training) |>
  step_log(all_numeric()) |> 
  step_normalize(all_numeric()) |> #Centering and scaling
  step_dummy(all_nominal(), -all_outcomes())
```

```{r Workflow object}
lr_worflow<-workflow() |> 
  add_model(lr_model) |> 
  add_recipe(lr_recipe)
```

```{r Fit the model}
set.seed(123)
lr_worflow_fit<-lr_worflow |> 
  last_fit(split = ml_split)

lr_worflow_fit |> 
  collect_metrics()

lr_resultss<-lr_worflow_fit |> 
  collect_predictions()

lr_resultss
```

## Logistic Regression Results

### Confusion Matrix

In this section we are going to visualize the model results. We will start with the **confusion matrix**.

```{r Confusion matrix}
set.seed(123)
lr_resultss |> 
  conf_mat(truth = diabetes,
           estimate = .pred_class)
```

-   The logistic regression correctly classifies 1,237 out of 1,358 individuals (91%). 116 false negatives and 8 false positives.

    ### Performance Metrics

```{r Metric set}
set.seed(112)
lr_metric<-metric_set(sens,accuracy, yardstick::spec)



lr_resultss |> 
  lr_metric(truth = diabetes, 
            estimate = .pred_class)

```

```{r Model metrics}
set.seed(123)
lr_resultss |>
  roc_curve(truth = diabetes,.pred_Yes) |>
  autoplot()


roc_auc(lr_resultss, truth = diabetes, .pred_Yes)

```

```{r}
set.seed(123)
heatmap_lr<-conf_mat(lr_resultss, truth = diabetes, estimate = .pred_class) |> 
  autoplot(type = "heatmap")


mosaic_lr<-conf_mat(lr_resultss, truth = diabetes, estimate = .pred_class) |> 
  autoplot(type = "mosaic")

cowplot::plot_grid(mosaic_lr,heatmap_lr)
```

The 91.1% accuracy is misleading. The model predicts negative cases well (high specificity) but identifies only 4 out of 120 positive cases. Accuracy alone is not a useful metric here due to class imbalance.

```{r}
set.seed(123)
roc_auc(lr_resultss,truth = diabetes,.pred_Yes)

lr_roc_plot<-lr_resultss |> 
  roc_curve(truth = diabetes, .pred_Yes) |> 
  autoplot()
  
lr_roc_plot
```

### Random Forest

Logistic regression failed to correctly identify positive cases. Random Forest offers more flexibility for nonlinear relationships and imbalanced data.

### Model Specification

```{r RF Model specification}
rf<-rand_forest() |> 
  set_args(mtry = 9) |> 
  set_engine("ranger", importance = "impurity") |> 
  set_mode("classification")
```

```{r Recipe}
set.seed(123)
rf_recipe<-recipe(diabetes~.,data = ml_training) |> 
  step_log(all_numeric()) |> 
  step_normalize(all_numeric()) |> 
  step_dummy(all_nominal(),-all_outcomes())
```

```{r Creating a  workflow object}
rf_workflow<-workflow() |> 
  add_model(rf) |> 
  add_recipe(rf_recipe)
```

```{r Fitting the model}
set.seed(123)
rf_wrkflw_fit<-rf_workflow |> 
  tune::last_fit(split = ml_split)

rf_wrkflw_fit |> 
  collect_metrics()

rf_results<-rf_wrkflw_fit |> 
  collect_predictions()
```

### Random Forest Results

```{r Conf Matrix}
set.seed(123)
mosaic_rf<-rf_results |> 
  conf_mat(truth = diabetes,.pred_class) |> 
  autoplot(type = "mosaic")

heatmap_rf<-rf_results |> 
  conf_mat(truth = diabetes,.pred_class) |>
  autoplot(type = "heatmap")
  
cowplot::plot_grid(heatmap_rf,mosaic_rf)
```

```{r Custom metrics}
set.seed(123)
rf_metrics<-metric_set(yardstick::sens, yardstick::spec, yardstick::accuracy)


lr_resultss |> 
  rf_metrics(truth = diabetes, 
            estimate = .pred_class)

rf_results |> 
  rf_metrics(truth = diabetes, estimate = .pred_class)

rf_results |> 
  roc_auc(truth = diabetes,.pred_Yes)
```

```{r ROC CURVE}
rf_roc_plot<-rf_results |> 
  roc_curve(truth = diabetes,.pred_Yes) |> 
  autoplot()
```

### Model comparison

```{r Model comparison}

cowplot::plot_grid(rf_roc_plot, lr_roc_plot,labels = c("Random Forest", "Logistic Regression"),label_size = 12)
```

Random Forest outperforms logistic regression on all metrics:

| Metric      | Logistic Regression | Random Forest |
|-------------|---------------------|---------------|
| Accuracy    | 0.91                | 0.93          |
| Sensitivity | 0.04                | 0.35          |
| Specificity | 0.99                | 0.99          |
| ROC-AUC     | 0.83                | 0.89          |

: The Random Forest improves sensitivity from 4% to 34%. The model still misses 64% of positive cases due to class imbalance.

### Prediction Example (Use case)

```{r}
patient1<-tribble(~age,~bmi,~direct_chol,~marital_status,~gender,
                  40,27.3,1.5,"Divorced","female")

rf_pred<-rf_wrkflw_fit |> 
  extract_workflow()

predict(rf_pred, new_data = patient1)
```

The model predicts this patient does not have diabetes. Given the model's high specificity, negative predictions are reliable.

## Limitations and Next Steps

This analysis skipped hyperparameter tuning (k-fold cross validation) and did not address class imbalance through oversampling (SMOTE) or threshold adjustment. These are the clear next steps for improving sensitivity on the minority class.
